Any way the brain blows? The nature of decision noise in random exploration

### 2022-12-21 OLD/oldversions/Wilson_Siyu_RandomDeterministicNoise/draft/Older drafts/Internal-External Noise/Internal-External Noise.tex
add older version

	\begin{figure}[hp]
		\begin{center}
			\includegraphics[width=\textwidth]{figures/taskfiga.pdf}
			\caption{BOB: ADD A PANEL A ABOVE THIS PANEL, SHOWING THE DYNAMICS FOR ONE GAME - LIKE IN HASHEM'S PAPER.  
			Schematic of the experiment. (A) Dynamics of an example horizon 6 game.  Here the first four trials are forced trials in which participants are instructed which option to play.  After the forced trials, participants are free to choose between the two options for the remainder of the game.  (B) Different possible states of the game after the first free choice over the course of the experiment. Overall participants play 160 such games, with varying horizon (1 vs 6), uncertainty condition ([1 3] vs [2 2]) and observed rewards.  In addition, all games are repeated (as Game 18 and 100 are here) such that participants will be faced with the exact same pattern of forced trials and exact same outcomes from those forced trials twice within each experiment.  These repeated games allow us to compute the relative contribution of external and internal noise by analyzing the extent to which choices are {\em consistent} across the repeated games.}%BOB: UP THE FONT SIZE FOR TEXT TO MATCH FONT SIZE FOR NUMBERS}
			\label{fig:taskfig}
		\end{center}
	\end{figure}

@@

	Random exploration is quantified in a model-free way as the probability of choosing the low mean option, $p(\mbox{low mean})$ in the equal, or [2 2], condition. This value increases with horizon, consistent with the idea that behavior is more random in horizon 6 (t(59) = 6.17, p $<$ 0.001 for [1 3], t(59) = 7.26, p $<$ 0.001 for [2 2]).  Directed exploration, is measured as the probability of choosing the more informative option $p(\mbox{high info}$ in the unequal, or [1 3], condition. Again this measure increases with horizon, showing that people are more information seeking in horizon 6 (t(59)=6.88, p $<$ 0.001).

@@

	\begin{figure}[H]
		\begin{center}
			\includegraphics[width=\textwidth]{figures/line_modelfree.png}
			\caption{BOB: DO THE YTICKS ON THE 0.1S NOT 0.15S!  MORE WHITE SPACE BETWEEN PANELS A AND B.
			Replication of previous findings. Both  $p(\mbox{low mean})$ (A) and $p(\mbox{high info})$ (B) increase with horizon suggesting that people use both random and directed exploration in this task.  }%BOB: SWITCH A AND B TO HAVE RANDOM FIRST}
			\label{fig:modelfree}
		\end{center}
	\end{figure}

@@

	To quantify choice inconsistency we computed the frequency with which participants made different responses for pairs of repeated games (Figure \ref{fig:mf2}). Using this measure we found that participants made inconsistent choices in both the unequal ([1 3]) and equal ([2 2]) information conditions (For [1 3] condition, t(59) = 11.78, p $<$ 0.001 for horizon 1, t(59) = 15.41, p $<$ 0.001 for horizon 6; For [2 2] condition, t(59) = 8.76, p $<$ 0.001 for horizon 1, t(59) = 17.02, p $<$ 0.001 for horizon 6.), suggesting that not all of the noise was stimulus driven. In addition, we found that choice inconsistency was higher in horizon 6 than in horizon 1 for both [1 3] and [2 2] condition (F(1, 236) = 37.98 BOB DID YOU FIX THIS? DIDNT USE REPEATED ANOVA, WILL CHANGE, p $<$ 0.001), suggesting that at least some of the horizon dependent noise is internal.

@@

	To gain more quantitative insight into these results, we computed theoretical values for the choice inconsistency for the purely external and purely internal noise cases.  For purely external noise this computation is simple because people should make the exact same decisions each time in repeated games, meaning that $p(\mbox{inconsistent}) = 0$ in this case. For purely internal noise, the two games should be treated independently, allowing us to compute the choice inconsistency in terms of the probability of choosing the low mean option, $p(\mbox{low mean})$, as
	\begin{equation*}
	\begin{split}
	p(\mbox{consistent}) &= p(\mbox{low mean})^2 + p(\mbox{high mean})^2\\
	&= p(\mbox{low mean})^2 + (1-p(\mbox{low mean}))^2\\ 
	\mbox{hence},\quad p(\mbox{inconsistent}) &=  2 p(\mbox{low mean})(1-p(\mbox{low mean}))
	\end{split}
	\end{equation*}
	As shown in Figure  \ref{fig:mf2}, people's behavior falls in between the pure external noise prediction and the pure internal noise prediction (Behavior is different from pure internal noise prediction in both [1 3] condition, t(59) = 5.10, p $<$ 0.001 for horizon 1, t(59) = 5.65, p $<$ 0.001 for horizon 6; and [2 2] condition, t(59) = 3.49, p $<$ 0.001 for horizon 1, t(59) = 3.31, p $<$ 0.001 for horizon 6. Behavior is different from pure external noise prediction in both [1 3] condition, t(59) = 11.78, p $<$ 0.001 for horizon 1, t(59) = 15.41, p $<$ 0.001 for horizon 6; and [2 2] condition, t(59) = 8.76, p $<$ 0.001 for horizon 1, t(59) = 17.02, p $<$ 0.001 for horizon 6.), suggesting that both external and internal noise are present in driving this choice inconsistency. Since choice inconsistency only reflects internal noise, Figure \ref{fig:mf2} suggests that internal noise increases with horizon.

@@

	80 participants (ages 18-25, 37 male, 43 female) from the University of Arizona undergraduate subject pool participated in the experiment. 20 were excluded on the basis of performance, using the same exclusion criterion as in \citep{wilson2014}. This left 60 for the main analysis. Note that including the 20 badly performing subjects did not change the main results (Supplementary Figures 1 - 3)% MAKE AND INCLUDE REPLICAS OF ALL FIGS BUT WITH ALL SUBJECTS

@@

	Each game consisted of 5 or 10 choices. Every game started with a fixation cross, then a bar of boxes will show up indicating the horizon for that game. For the first 4 games - the instructed games, we highlight the box on one of the bandits to instruct the participant to choose that option, they have to press the corresponding key to reveal the outcome. From the $5^{th}$ trial, boxes on both bandits will be highlighted and they are free to make their own decision. There was no time limit for decisions. During free choices they could press either the left arrow key or right arrow key to indicate their choice of left or right bandit. The score feedback was presented for 300ms. The task was programmed using Psychtoolbox in MATLAB \citep{psychtoolbox1, psychtoolbox2}. (See Figure \ref{fig:taskfigb})

	The first four trials of each game were forced-choice trials, in which only one of the options was available for the participant to choose. We used these forced-choice trials to manipulate the relative ambiguity of the two options, by providing the participant with different amounts of information about each bandit before their first free choice. The four forced-choice trials set up two uncertainty conditions: unequal uncertainty(or [1 3]) in which one option was forced to be played once and the other three times, and equal uncertainty(or [2 2]) in which each option was forced to be played twice. After the forced-choice trials, participants made either 1 or 6 free choices (two horizon conditions),  Figure \ref{fig:taskfigb}.

	\begin{figure}[H]
		\begin{center}
			\includegraphics[width=\textwidth]{figures/taskfigb.pdf}
			\caption{BOB: DONT NEED THIS FIGURE IF YOU UPDATE EARLIER TASKFIGURE. Time line of a horizon 6 game.  The game begins with four forced choice trials in which participants are instructed which option to play.   After these forced trials they are free to choose between the two options for the rest of the game (6 trials in this case).
			%BOB: THIS FONT SIZE IS GOOD - MAYBE A LITTLE BIGGER FOR TEXT
			%COMBINE THIS WITH FIGURE 7 FOR TASK CONDITIONS
			%CAN YOU MAKE THESE FIGURES LOOK NICER I.E. MORE LIKE MINE?  REDUCE LINE WIDTHS ETC ... IF NOT SEND ME POWER POINTS AND I WILL MAKE INTO PDFS}
			%I WILL SEND YOU THE POWERPOINTS TOO
			}
			\label{fig:taskfigb}
		\end{center}
	\end{figure}
%	\begin{figure}[H]
%		\begin{center}
%			\includegraphics[width=\textwidth]{figures/taskfig2.PNG}
%			\caption{Task conditions}
%			\label{fig:taskfig2}
%		\end{center}
%	\end{figure}

@@

	\begin{figure}[H]
		\begin{center}
			\includegraphics[width=1\textwidth]{figures/Siyu_EEHorizon_2sigma_final.pdf}
			\caption{Schematic of the hierarchical Bayesian model using notation of BOB CITE WAGNERMAKERS AND LEE.}
			\label{fig:model}
		\end{center}
	\end{figure}

>>>>

	\begin{figure}[hp]
		\begin{center}
			\includegraphics[width=\textwidth]{figures/taskfiga.pdf}
			\caption{
			%BOB: ADD A PANEL A ABOVE THIS PANEL, SHOWING THE DYNAMICS FOR ONE GAME - LIKE IN HASHEM'S PAPER.  
			Schematic of the experiment. (A) Dynamics of an example horizon 6 game.  Here the first four trials are forced trials in which participants are instructed which option to play.  After the forced trials, participants are free to choose between the two options for the remainder of the game.  (B) Different possible states of the game after the first free choice over the course of the experiment. Overall participants play 160 such games, with varying horizon (1 vs 6), uncertainty condition ([1 3] vs [2 2]) and observed rewards.  In addition, all games are repeated (as Game 18 and 100 are here) such that participants will be faced with the exact same pattern of forced trials and exact same outcomes from those forced trials twice within each experiment.  These repeated games allow us to compute the relative contribution of external and internal noise by analyzing the extent to which choices are {\em consistent} across the repeated games.}%BOB: UP THE FONT SIZE FOR TEXT TO MATCH FONT SIZE FOR NUMBERS}
			\label{fig:taskfig}
		\end{center}
	\end{figure}

@@

	Random exploration is quantified in a model-free way as the probability of choosing the low mean option, $p(\mbox{low mean})$ in the equal, or [2 2], condition. This value increases with horizon, consistent with the idea that behavior is more random in horizon 6 (t(65) = 6.60, p $<$ 0.001 for [1 3], t(65) = 8.14, p $<$ 0.001 for [2 2]).  Directed exploration, is measured as the probability of choosing the more informative option $p(\mbox{high info}$ in the unequal, or [1 3], condition. Again this measure increases with horizon, showing that people are more information seeking in horizon 6 (t(65) = 7.00, p $<$ 0.001).

@@

	\begin{figure}[H]
		\begin{center}
			\includegraphics[width=\textwidth]{figures/line_modelfree.png}
			\caption{
			%BOB: DO THE YTICKS ON THE 0.1S NOT 0.15S!  MORE WHITE SPACE BETWEEN PANELS A AND B.
			Replication of previous findings. Both  $p(\mbox{low mean})$ (A) and $p(\mbox{high info})$ (B) increase with horizon suggesting that people use both random and directed exploration in this task.  }%BOB: SWITCH A AND B TO HAVE RANDOM FIRST}
			\label{fig:modelfree}
		\end{center}
	\end{figure}

@@

	To quantify choice inconsistency we computed the frequency with which participants made different responses for pairs of repeated games (Figure \ref{fig:mf2}). Using this measure we found that participants made inconsistent choices in both the unequal ([1 3]) and equal ([2 2]) information conditions (For [1 3] condition, t(65) = 12.92, p $<$ 0.001 for horizon 1, t(65) = 16.76, p $<$ 0.001 for horizon 6; For [2 2] condition, t(65) = 9.67, p $<$ 0.001 for horizon 1, t(65) = 17.74, p $<$ 0.001 for horizon 6.), suggesting that not all of the noise was stimulus driven. In addition, we found that choice inconsistency was higher in horizon 6 than in horizon 1 for both [1 3] and [2 2] condition (F(1, 196) = 61.19, p $<$ 0.001), suggesting that at least some of the horizon dependent noise is internal.

@@

	To gain more quantitative insight into these results, we computed theoretical values for the choice inconsistency for the purely external and purely internal noise cases.  For purely external noise this computation is simple because people should make the exact same decisions each time in repeated games, meaning that $p(\mbox{inconsistent}) = 0$ in this case. For purely internal noise, the two games should be treated independently, allowing us to compute the choice inconsistency in terms of the probability of choosing the low mean option, $p(\mbox{low mean})$, as
	\begin{equation*}
	\begin{split}
	p(\mbox{consistent}) &= p(\mbox{low mean})^2 + p(\mbox{high mean})^2\\
	&= p(\mbox{low mean})^2 + (1-p(\mbox{low mean}))^2\\ 
	\mbox{hence},\quad p(\mbox{inconsistent}) &=  2 p(\mbox{low mean})(1-p(\mbox{low mean}))
	\end{split}
	\end{equation*}
	As shown in Figure  \ref{fig:mf2}, people's behavior falls in between the pure external noise prediction and the pure internal noise prediction (Behavior is different from pure internal noise prediction in both [1 3] condition, t(65) = 5.60, p $<$ 0.001 for horizon 1, t(65) = 5.62 p $<$ 0.001 for horizon 6; and [2 2] condition, t(65) = 3.86, p $<$ 0.001 for horizon 1, t(65) = 3.42, p $<$ 0.001 for horizon 6. Behavior is different from pure external noise prediction in both [1 3] condition, t(65) = 12.92, p $<$ 0.001 for horizon 1, t(65) = 16.76, p $<$ 0.001 for horizon 6; and [2 2] condition, t(65) = 9.67, p $<$ 0.001 for horizon 1, t(65) = 17.74, p $<$ 0.001 for horizon 6.), suggesting that both external and internal noise are present in driving this choice inconsistency. Since choice inconsistency only reflects internal noise, Figure \ref{fig:mf2} suggests that internal noise increases with horizon.

@@

	80 participants (ages 18-25, 37 male, 43 female) from the University of Arizona undergraduate subject pool participated in the experiment. 14 were excluded on the basis of performance, using the same exclusion criterion as in \citep{wilson2014}. This left 66 for the main analysis. Note that including the 14 badly performing subjects did not change the main results (Supplementary Figures 1 - 3)% MAKE AND INCLUDE REPLICAS OF ALL FIGS BUT WITH ALL SUBJECTS

@@

	Each game consisted of 5 or 10 choices. Every game started with a fixation cross, then a bar of boxes will show up indicating the horizon for that game. For the first 4 games - the instructed games, we highlight the box on one of the bandits to instruct the participant to choose that option, they have to press the corresponding key to reveal the outcome. From the $5^{th}$ trial, boxes on both bandits will be highlighted and they are free to make their own decision. There was no time limit for decisions. During free choices they could press either the left arrow key or right arrow key to indicate their choice of left or right bandit. The score feedback was presented for 300ms. The task was programmed using Psychtoolbox in MATLAB \citep{psychtoolbox1, psychtoolbox2}. (See Figure \ref{fig:taskfig})

	The first four trials of each game were forced-choice trials, in which only one of the options was available for the participant to choose. We used these forced-choice trials to manipulate the relative ambiguity of the two options, by providing the participant with different amounts of information about each bandit before their first free choice. The four forced-choice trials set up two uncertainty conditions: unequal uncertainty(or [1 3]) in which one option was forced to be played once and the other three times, and equal uncertainty(or [2 2]) in which each option was forced to be played twice. After the forced-choice trials, participants made either 1 or 6 free choices (two horizon conditions),  Figure \ref{fig:taskfig}.

%	\begin{figure}[H]
%		\begin{center}
%			\includegraphics[width=\textwidth]{figures/taskfigb_final.pdf}
%			\caption{BOB: DONT NEED THIS FIGURE IF YOU UPDATE EARLIER TASKFIGURE. Time line of a horizon 6 game.  The game begins with four forced choice trials in which participants are instructed which option to play.   After these forced trials they are free to choose between the two options for the rest of the game (6 trials in this case).
			%BOB: THIS FONT SIZE IS GOOD - MAYBE A LITTLE BIGGER FOR TEXT
			%COMBINE THIS WITH FIGURE 7 FOR TASK CONDITIONS
			%CAN YOU MAKE THESE FIGURES LOOK NICER I.E. MORE LIKE MINE?  REDUCE LINE WIDTHS ETC ... IF NOT SEND ME POWER POINTS AND I WILL MAKE INTO PDFS}
			%I WILL SEND YOU THE POWERPOINTS TOO
%			}
%			\label{fig:taskfigb}
%		\end{center}
%	\end{figure}
%	\begin{figure}[H]
%		\begin{center}
%			\includegraphics[width=\textwidth]{figures/taskfig2.PNG}
%			\caption{Task conditions}
%			\label{fig:taskfig2}
%		\end{center}
%	\end{figure}

@@

	\begin{figure}[H]
		\begin{center}
			\includegraphics[width=1\textwidth]{figures/Siyu_EEHorizon_2sigma_final.pdf}
    			\caption{Schematic of the hierarchical Bayesian model using notation of \cite{lee2014}}
			%BOB CITE WAGNERMAKERS AND LEE.}
			\label{fig:model}
		\end{center}
	\end{figure}
